Split-Apply-Combine with Collapsing Groups

### 2022-12-30 paper/article.Rnw
writing

Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented
in practice. Consider the signature of the R \texttt{aggregate} function for data
frames (we skip arguments that are not important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \texttt{x} is a data frame where each column is a variable $U$ to be
aggregated.  The argument \texttt{by} represents the record-wise label(s) that
are used for grouping. So the function $f\to A$ is implemented by forcing the
user to make sure that the position of each label in \texttt{by} corresponds to
the row number in the data frame \texttt{x}. The argument \texttt{FUN} is the
function $\phi$ that aggregates each subset of \texttt{x}.  Here is an example of how
a user might call this function from the R prompt.
%
\begin{verbatim}
> aggregate(iris["Sepal.Length"], by=iris["Species"], FUN=mean)
     Species Sepal.Length
1     setosa        5.006
2 versicolor        5.936
3  virginica        6.588
\end{verbatim}
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The output
also reveals in the first column the set $A$: each row corresponds to a unique
value in \texttt{Species} column.

@@

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{table}[h!]
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\label{eq:tablerep}
\end{table}

@@

\subsection{Caveat: unbalanced hierarchical classifications}
\begin{itemize}
\item tree with different levels of details
\end{itemize}

\section{The accumulate R package}\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. The main function of the package is caled \texttt{collapse}
and has the following signature.
\begin{verbatim}
  accumulate(data, collapse, test, ...)
\end{verbatim}
where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
of \texttt{label = expression} pairs that represent aggregation
functions, in a similar style as \texttt{dplyr::summarize}. The output
is a data frame that can schematically be represented as follows.
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate, and the last set of columns store the aggregates.

To accommodate the different use cases, the \pkg{accumulate} package supports
two different interfaces for specifying collapse sequences.  The first and most
general is the formula interface, that requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is the
tabular interface discussed before. This only supports a single grouping
variable. For hierarchical classifications there is a helper function to
translate parent-child representations of such classifications into a
collapsing sequence.

Besides complicated collapsing schemes, users must have the ability to specify
a possibly complex condition for determining when a dataset is suitable for
aggregation or not. The package offers some convenience functions that support common
use cases as well as support for specifying generic conditions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
It is not uncommon for users to use multiple variables for defining collapsing
schemes. For example when grouping by two categorical variables \texttt{A} and
\texttt{B}, one may first want to collapse variable $\texttt{B}$ into a
variable \texttt{B1} and if that is not sufficient, in the next step grouping
in the second variable is dropped so grouping is only defined by \texttt{A}.
In \pkg{accumulate} such a scheme is denoted with the formula \texttt{A*B \~{}
A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
that is attempted first, and the right-hand-side denotes consecutive collapse
levels separated by \texttt{+}.

The following tables illustrate the idea. The left table represents the data
set to be aggregated by \texttt{A} and \texttt{B}. In this example we wish to
compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
least three records in a group, otherwise we collapse one step.
%
\begin{center}
\begin{tabular}{cccl}
\multicolumn{4}{l}{\texttt{Input data}}\\
\hline
\texttt{A} & \texttt{B} & \texttt{B1} & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{\texttt{Output}}\\
\hline
\texttt{A*B} & \texttt{level} & $t_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 7  \\
\rowcolor{green!25} 2 12        & 1             & 56 \\
\rowcolor{green!25} 2 13        & 1             & 56 \\
\rowcolor{blue!25}  3 21        & 2             & 448\\
\rowcolor{blue!25}  3 22        & 2             & 448\\
\rowcolor{blue!25}  3 12        & 2             & 448\\
\hline
\end{tabular}
\end{center}
%
The table on the right represents the output. Colors indicate which data was
used. The First row in the output represents group
$(\texttt{A}=1,\texttt{B}=11)$. The collapsing level is zero, which means
that no collapsing was necessary. Indeed, in the data table we see that there
are three rows with \texttt{A==1} and \texttt{B==11} with $Y$ values $1, 2$,
and $7$, resulting in $t_Y=7$ for this group.

Next, the algorithm tries to compute the total for group
$(\texttt{A}=2,\texttt{B}=12)$ (in green) but finds that there are only two
such rows. It now defines a new group $(\texttt{A}=2,\texttt{B1}=1)$ and finds
that there are three records in that group so it computes $t_Y=8+16+32=56$.
Similarly, it finds one record with $(\texttt{A}=2,\texttt{B}=13)$ and
collapses to $(\texttt{A}=2,\texttt{B1}=1)$, again yielding $t_y=56$.

Finally, for $(\texttt{A}=2,\texttt{B}=21)$ the algorithm finds only a single
record. It collapses to $(\texttt{A}=2,\texttt{B1}=2)$ and finds two so it
collapses further to $(\texttt{A}=2)$ to find three records.  This yields
$t_Y=64+128+256=448$. Similarly, the groups $(\texttt{A}=3, \texttt{B}=22)$ and
$(\texttt{A}=3,\texttt{B}=12)$ are collapsed to $(\texttt{A}=2)$.

In R code the previous example can be executed as follows.
\begin{verbatim}
  > library(accumulate)
  > input <- data.frame(
  +      A  = c(1,1,1,2,2,2,3,3,3)
  +    , B  = c(11,11,11,12,12,13,21,22,12)
  +    , B1 = c(1,1,1,1,1,1,2,2,1)
  +    , Y  = 2^(0:8)
  + )
  >
  > collapse(data     = input
  +        , collapse = A*B ~ A*B1 + A
  +        , test     = function(d) nrow(d)>=3
  +        , tY       = sum(Y) )
    A  B level  tY
  1 1 11     0   7
  2 2 12     1  56
  3 2 13     1  56
  4 3 21     2 448
  5 3 22     2 448
  6 3 12     2 448
\end{verbatim}
%
The formula interface allows users to quickly experiment with different
collapsing schemes, but only after the necessary variables have been added to
the original data frame, which may require some extensive data manipulation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The data frame interface}
\eqref{eq:tablerep}

>>>>

Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented
in practice. Consider the signature of the R \code{aggregate()} function for data
frames (we skip arguments that are not important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \code{x} is a data frame where each column is a variable $U$ to be
aggregated.  The parameter \code{by} represents the record-wise label(s) that
are used for grouping. So the function $f\to A$ is implemented by forcing the
user to make sure that the position of each label in \code{by} corresponds to
the row number in the data frame \code{x}. The argument \code{FUN} is the
function $\phi$ that aggregates each subset of \code{x}.  Here is an example of how
a user might call this function from the R prompt.
%
<<>>=
aggregate(iris[1:2], by=iris["Species"], FUN=mean)
@
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The output
also reveals in the first column the set $A$: each row corresponds to a unique
value in \texttt{Species} column.

@@

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{equation}
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\label{eq:tablerep}
\end{equation}

@@

\subsection{Performance, memory, paralellization}
Both Algorithms~\ref{alg:sac} and \ref{alg:saccg} need only $\mathcal{O}(|U|)$
memory. The time complexity of Algorithm~\ref{alg:sac} is $\mathcal{O}(|A||U|)$
($|A|$ the number of groups occurring in the data, and $|U|$ the number of
records) but the actual running time is also determined by the form of the
aggregating function $\phi$.  For simple aggregating functions such as
summation or counting, the output can be computed with a single pass through
$|U|$. \proglang{R} packages such as \pkg{collapse} and \pkg{data.table} have
implemented specialized low level (\proglang{C}/\proglang{C++}) code for fast
groupwise aggregation for some commonly occurring aggregating functions.
Generic paralellization of Algorithm~\ref{alg:sac} can be done by paralellising
over the grouping values $a\in A$: each worker node gets access to a set of
records corresponding to a single group and computes the local aggregate.

Algorithm~\ref{alg:saccg} has time complexity of $\mathcal{O}((n+1)|A_0||U|)$,
Namely, in the worst case, for each $a\in A_0$ we need to collapse to the
maximum level $A_n$. This means that for each $a\in A_0$ we need to collapse
$n$ times, plus the initial attempt of computing aggregate without collapsing.
The fact that we cannot generically exclude records from a calculation means
that a generic paralellisation scheme would need to ensure that each worker
node has access to the full dataset $|U|$. There are two types of options to do
this: either mutliple workers gain read-only access to $U$ or each worker gets
a copy of $U$. In the first case the workers may suffer from colliding
read-operations which hampers linear performance improvement as a function of
the number of workers. In the second case, one needs as many copies of the data
as there are threads. It is possible to compute beforehand whether the dataset
can be split into subsets that can be treated independently by determining all
the collapsed subsets. However, this is almost as expensive as running the
algorithm completely since it only skips the calculation of aggregates after
collapsing.

Summarizing, we see that split-apply-combine with collapsing groups has a
higher worst-case complexity than regular split-apply-combine. Moreover,
generic paralellization needs to either consider a specific use case (e.g.
large or small datasets) or require user input that guides the choice of
paralellisation scheme.

\section{The accumulate R package}\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. There are two main functions in \pkg{accumulate} that offer
slightly different interfaces.
\begin{verbatim}
  accumulate(data, collapse, test, fun, ...)
  cumulate(data, collapse, test, ...)
\end{verbatim}
Here \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence (as a \class{formula} or a \class{data frame}), and \texttt{test} is a
function that accepts a subset of \texttt{data} and returns a boolean that
indicates whether a subset is suited for aggregation. In \code{accumulate()},
the parameter \code{fun} represents an aggregation function that is applied to
every column of \code{data}, and the ellepsis (\code{...}) is for arguments
that are passed as extra argument to \code{fun}. The interface of
\code{accumulate()} is modeled after the \code{aggregate()} function in
\proglang{R}.  In \code{cumulate}, the ellepsis is a sequence of
comma-separated \code{name=expression} pairs in the style of \code{summarise()}
from the \code{dplyr} package.

The output of both functions are of the same form.  The columns of the output
data frame and can schematically be represented as follows. 
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate (0 indicating no collapse), and the last set of columns
store the aggregates.

Both functions support two different interfaces for specifying collapsing
schemes through the \code{collapse} parameter. The first and most general is
the formula interface, which requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is the
tabular interface, where a table similar to~\eqref{eq:tablerep} is expected.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
It is not uncommon for users to use multiple variables for defining collapsing
schemes. For example when grouping by two categorical variables $A$ and
$B$, one may first want to collapse variable $B$ into a
variable $B1$ and if that is not sufficient, grouping
in the second variable is dropped so grouping is only defined by $A$.
In \code{accumulate()} such a scheme is denoted with the formula \code{A*B \~{}
A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
that is attempted first, and the right-hand-side denotes consecutive collapse
levels separated by \code{+}.

The following tables illustrate the idea. The left table represents the data
set to be aggregated by $A$ and $B$. In this example we wish to
compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
least three records in a group, otherwise we collapse one step (we set values
of $Y$ to unique powers of two, so it is clear which records are used
to compute each value of $t_Y$).
%
\begin{center}
\begin{tabular}{cccl}
\multicolumn{4}{l}{Input data}\\
\hline
$A$ & $B$ & $B1$ & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{Output}\\
\hline
$A\times B$ & \code{level} & $t_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 7  \\
\rowcolor{green!25} 2 12        & 1             & 56 \\
\rowcolor{green!25} 2 13        & 1             & 56 \\
\rowcolor{blue!25}  3 21        & 2             & 448\\
\rowcolor{blue!25}  3 22        & 2             & 448\\
\rowcolor{blue!25}  3 12        & 2             & 448\\
\hline
\end{tabular}
\end{center}
%
The table on the right represents the output. Colors indicate which data was
used. The First row in the output represents group
$(A=1,B=11)$. The collapsing level is zero, which means
that no collapsing was necessary. Indeed, in the data table we see that there
are three rows with $A=1$ and $B==11$ with $Y$ values $1, 2$,
and $7$, resulting in $t_Y=7$ for this group.

Next, the algorithm tries to compute the total for group
$(A=2,B=12)$ (in green) but finds that there are only two
such rows. It now defines a new group $(A=2,B1=1)$ and finds
that there are three records in that group so it computes $t_Y=8+16+32=56$.
Similarly, it finds one record with $(A=2,B=13)$ and
collapses to $(A=2,B1=1)$, again yielding $t_y=56$.

Finally, for $(A=2,B=21)$ the algorithm finds only a single
record. It collapses to $(A=2,B1=2)$ and finds two so it
collapses further to $(A=2)$ to find three records.  This yields
$t_Y=64+128+256=448$. Similarly, the groups $(A=3, B=22)$ and
$(A=3,B=12)$ are collapsed to $(A=2)$.

In R code the previous example can be executed as follows.
<<>>=
library(accumulate)
input <- data.frame(
     A  = c(1,1,1,2,2,2,3,3,3)
   , B  = c(11,11,11,12,12,13,21,22,12)
   , B1 = c(1,1,1,1,1,1,2,2,1)
   , Y  = 2^(0:8)
)

cumulate(input, collapse = A*B ~ A*B1 + A
        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
@
%
The formula interface allows users to quickly experiment with different
collapsing schemes, but only after the necessary variables have been added to
the original data frame.

It is possible to get the same result with \code{accumulate()}. This will 
cause summation over all variables that are not used in the formula object. In
the below example we also introduce the helper function \code{min_records()}.
<<>>=
input$Y2 <- 3^(0:8)
accumulate(input, collapse = A*B ~ A*B1 + A
          , test = min_records(3), fun = sum)
@
This means that for experimentation, users must be careful to exclude categorical
variables that are not used from the input data set. For example, if \code{B1}
is not used, we get the following.
<<>>=
accumulate(input[-3], collapse = A*B ~ A
          , test = min_records(3), fun = sum)
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The data frame interface} \label{sect:dfint}
The data frame interface is somewhat limited because it only allows for a
single grouping variable. The advantage however setting up a collapsing scheme
in the form of a table closely connects to domain knowledge and allows
fine-grained control on how groups are collapsed.

In order to use the data frame interface, the input dataset needs to have the
most fine-grained grouping variable. In the running example this is
$A\times B$, so we need to combine that into a single variable, and remove
the other ones.
<<>>=
input$AB <- paste(input$A, input$B, sep="-")
input <- input[-(1:3)]
input
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
@
The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
<<>>=
accumulate(input, collapse=csh, test = min_records(3), sum)
cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
@

\subsection{Balanced and unbalanced hierarchical classifications}
